---
title: "Standard pre-processing of scRNA-seq analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this tutorial, we will be analyzing the Peripheral Blood Mononuclear Cells (PBMC) data available from 10X Genomics. The data comprises of 2,700 single cells that were sequenced on the Illumina NextSeq 500.

### 1. Set working directory and load libraries

```{r}
# setwd("Github/10XGenomics_Pipeline/1. scRNA-seq/")
```

```{r}
# Tool for QC, normalization, dimensionality reduction and clustering of single cell datsests
library(Seurat) 

## Tool for data manipulation, cleaning, and visualization of tabular data
library(tidyverse) 
```


### 2. Load the PBMC dataset

The raw data can be found here- https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz. Download and extract the tar.gz archive. This unpacks the barcode, gene, and matrix files needed for downstream analysis.

```{r}
# Read the 10x Genomics matrix directory into R
pbmc.data <- Read10X(data.dir = "./filtered_gene_bc_matrices/hg19/")

# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)

# Print a summary of the newly created Seurat object
pbmc
```

```{r}
# Lets examine a few genes in the first thirty cells
pbmc.data[c("CD3D", "TCL1A", "MS4A1"), 1:30]
```

The symbol "." represent zeros (no detected molecules). Because single-cell RNA-seq datasets are inherently "sparse"—meaning the vast majority of values are zero—Seurat stores this data as a sparse matrix. By only tracking the non-zero coordinates rather than storing every single zero, Seurat drastically reduces memory usage and speeds up processing times for droplet-based workflows (like 10x Genomics, Drop-seq, and inDrop).

#### 3. Standard pre-processing workflow

This pipeline takes noisy, high-dimensional raw sequencing counts and processes them into biologically interpretable clusters:

#### 3a. Quality Control (QC)

Droplet-based single-cell protocols are imperfect; they frequently capture empty droplets, cell doublets (two cells trapped together), or dying cells. To ensure our downstream analysis is built only on viable, high-quality cells, we evaluate three primary metrics:

**nFeature_RNA** : number of distinct genes detected in an individual cell. Low counts usually indicate empty droplets or poor-quality cells where the sequencing failed. High counts often indicate doublets or multiplets (two or more cells captured in a single droplet).

**nCount_RNA** : total number of unique molecular identifiers (UMIs) detected within a cell. This correlates strongly with the unique gene count and helps identify overall sequencing depth per cell.

**percent.mt** : proportion of reads that map to the mitochondrial genome. High percentages are a classic signature of dying cells.

```{r}
# Calculate the percentage of total reads that map to mitochondrial genes for each cell.
pbmc[["percent.mt"]] <- PercentageFeatureSet(object = pbmc, pattern = "^MT-")

# Show QC metrics for the first few rows of the metadata
head(pbmc[])
```

```{r}
# Visualize QC metrics as a violin plot
VlnPlot(object = pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

**nFeature_RNA (Unique Genes)**: Most cells contain between 500 and 1,000 unique genes. The long thin tail at the top shows a few cells reaching over 2,500 to 3,000 genes—these are highly likely to be doublets (two cells in one droplet) and should be trimmed.

**nCount_RNA (Total Molecules/UMIs)**: This mirrors the gene count plot, showing that most cells have around 2,500 total RNA molecules detected, with a few extreme outliers extending up to 15,000.

**percent.mt (Mitochondrial Percentage)**: The vast majority of your cells have a very low mitochondrial read percentage (under 5%), which indicates a highly viable sample. However, you can see a distinct spike of dying cells extending all the way up to 20–25% mitochondrial content.

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships
plot1 <- FeatureScatter(object = pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(object = pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
CombinePlots(plots = list(plot1, plot2))
```

**nCount_RNA vs. percent.mt**: High-quality, healthy cells (low percent.mt shown in the bottom-left cluster) generally scale up to have a healthy amount of total RNA (nCount_RNA). Conversely, the dead or dying cells sitting near the top of the y-axis (>10% mitochondrial reads) have very low overall counts because their cytoplasmic RNA has leaked away.

**nCount_RNA vs. nFeature_RNA**: Near-perfect linear correlation of 0.95 is the hallmark of a high-quality dataset, more molecules from a cell correlate to more unique genes. Cells that lie way up and to the right are suspect for being doublets. Cells clustered near the bottom-left are typically empty droplets or low-quality cells that failed to capture sufficient RNA.

```{r}
# Filter the dataset to retain only high-quality cells based on QC thresholds.
pbmc <- subset(x = pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

# Print a summary of the newly filtered Seurat object
pbmc
```

Filtering step successfully discarded 62 low-quality cells (dropping from 2,700 to 2,638 samples) that failed the gene count or mitochondrial thresholds. The number of genes remained exactly the same (13,714), as the subset function only removes poor-quality cells (columns) without altering the genomic features (rows).

#### 3b. Data Normalization

Different cells are sequenced to different depths, meaning some cells naturally have more total reads than others purely due to technical variation.

The traditional workflow uses **LogNormalize, FindVariableFeatures followd by ScaleData** which often fails to completely remove the technical noise associated with sequencing depth. Highly sequenced cells still cluster together purely because they have more reads.

**SCTransform** (introduced by Hafemeister and Satija) replaces those three distinct steps with a single, elegant framework based on regularized negative binomial regression.

```{r}
# Run ScTransform normalization and scaling while regressing technical variation caused by mitochondrial contamination/cell stress
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
```

#### 3c. Perform linear dimensional reduction

```{r}
# Perform Principal Component Analysis (PCA) on the Seurat object
pbmc <- RunPCA(pbmc, verbose = FALSE)
```

```{r}
# Print the top-contributing features (genes) for the first 5 Principal Components.
print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
```

**Cell Type Signatures**: 

**PC_1**: Driven on the positive end by myeloid/monocyte markers (LYZ, CST3, S100A9). 

**PC_2**: Separates B cells/Antigen-Presenting Cells (positive markers: MHC-II genes like HLA-DRA, CD74, CD79A) from Cytotoxic T and NK cells (negative markers: killer proteins like NKG7, GZMB, GNLY, GZMA). 

**PC_4**: Distinctly separates CD16+ non-classical monocytes (positive markers: FCGR3A / CD16, LST1) from classical CD14+ monocytes (negative markers: CD14, S100A8, S100A9). 

**PC_5**: Captures Megakaryocytes/Platelets on its negative axis (markers: PPBP, PF4). 

```{r}
# Generate a heatmap for the first Principal Component
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

**Left-side Cells (Myeloid/Monocytes)**: high expression of S100A8, S100A9, LST1, FTL, and FTH1. 

**Right-side Cells (Lymphocytes - T, B, and NK cells)**: high expression of ribosomal proteins (RPS/RPL genes), LTB (Lymphotoxin-beta), and NKG7 (Natural killer cell granule protein 7). 

```{r}
# Determine dimensionality of the data
ElbowPlot(pbmc)
```
For this dataset, selecting 10 dimensions (dims = 1:10) is the classic, ideal cutoff. This threshold successfully captures over 90% of the true biological variation while safely excluding the flat "tail" of random background noise.

#### 3d. Clustering

Seurat first construct a KNN graph based on the euclidean distance in the 10-dimensional PCA space, and refine the edge weights between any two cells based on the shared overlap in their local neighborhoods (Jaccard similarity).

```{r}
# Construct Shared Nearest-neighbor (SNN) graph 
pbmc <- FindNeighbors(pbmc, dims = 1:10, verbose = FALSE)
```

```{r}
# Cluster the cells using a modularity optimization algorithm (like the Louvain algorithm).
# The resolution parameter that sets the granularity of the downstream clustering, with increased values leading to a greater number of clusters
pbmc <- FindClusters(pbmc, verbose = FALSE, resolution = c(0.2, 0.4, 0.6, 0.8, 1))
```

```{r}
# # Visualize the impact of different clustering granularities (resolutions) on your UMAP:
DimPlot(pbmc, group.by = "SCT_snn_res.0.2", label = TRUE)
DimPlot(pbmc, group.by = "SCT_snn_res.0.4", label = TRUE)
DimPlot(pbmc, group.by = "SCT_snn_res.0.6", label = TRUE)
DimPlot(pbmc, group.by = "SCT_snn_res.0.8", label = TRUE)
```

These plots project cells onto PC_1 vs. PC_2 rather than a UMAP. Because PCA is a linear reduction, looking at only the first two dimensions causes different cell types to overlap and look stacked on top of each other.

```{r}
# Run the Uniform Manifold Approximation and Projection (UMAP) dimensional reduction.
pbmc <- RunUMAP(object = pbmc, dims = 1:10)
```

```{r}
# Plot UMAP colored by a specific clustering resolution
DimPlot(pbmc, reduction = "umap", group.by = "SCT_snn_res.0.2")
DimPlot(pbmc, reduction = "umap", group.by = "SCT_snn_res.0.4")
DimPlot(pbmc, reduction = "umap", group.by = "SCT_snn_res.0.6")
DimPlot(pbmc, reduction = "umap", group.by = "SCT_snn_res.0.8")
```

Once you run UMAP—which is non-linear and uses your cell-to-cell neighbor graph—these overlapping groups will stretch and untangle into beautifully isolated "islands" representing distinct cell types.

* **res.0.2 (Under-clustered)**: Too broad. 
* **res.0.4 (The Sweet Spot)**: Best for most analyses. It cleanly separates major cell subtypes (like splitting the top-right T-cell island into two halves) without creating messy, artificial boundaries.
* **res.0.6 (Highly Detailed)**: Good if you want to study subtle functional states or find rare immune cell populations.
* **res.0.8 (Over-clustered)**: Too granular. The speckled, intermingled clusters on the top-right island show that the algorithm is drawing arbitrary lines through a single, continuous cell population.

```{r}
# Setting identity of clusters
Idents(pbmc) <- "SCT_snn_res.0.4"
View(pbmc@meta.data)
```

### QC Additions to Strengthen the Pipeline

**1. Ambient RNA Removal (DeconTx)**
Immediately after creating the Seurat object (before any filtering

**2. Doublet Removal (DoubletFinder)**
Directly after running your initial PCA and UMAP, but before final clustering.

**3. Cluster Resolution Selection (clustree)**
Right after running FindClusters with multiple resolutions.

### Downstream Analysis (Next Steps)

**1. Marker Gene Identification & Cell-Type Annotation**
To discover what cell types your UMAP islands actually represent, run differential expression to find highly enriched genes in each cluster and then cross-reference these marker genes with known biological databases.

**2. Differential Expression (DE) & Pathway Analysis**
Compare gene expression profiles between conditions (e.g., Healthy vs. Treated CD4+ T-cells) to identify upregulated pathways using tools like Gene Ontology (GO) or GSEA (Gene Set Enrichment Analysis).

**3. Trajectory Inference (Pseudotime Analysis)**
Model cells along a transitional timeline using tools like Monocle3, useful for studying continuous biological processes, such as stem cell differentiation or T-cell exhaustion